### 2021-09-06 ch/3-structure.tex
Final version
Fix small errors found during corrections

The described method works particularly well with \acp{INN},
as they ensure that no information is lost that is unrelated to \(s\).
However, the price to pay for using \acp{INN} is higher memory requirement and slower training.
Thus, we also present a variant of the method using \iac{VAE},
which does not have the guarantee about preserving information,
but also does not suffer from the increased training cost as much.
\acp{VAE} are similar to \acp{INN}
in that their encoding conforms to a specific probability distribution,
from which we can sample our null-samples.
The choice between the two presented variants is determined
by whether the user is willing to accept higher training costs
for a lower probability of losing information needed for any prediction tasks.
However, the key element is simply any kind encoder -- producing a split-encoding --
whose output can be subjected to adversarial training,
and other kinds of encoders will potentially work as well.

@@

The absolutely essential elements for this method are the encoder (also referred to as `de-biaser')
that encodes both, samples from the training set and samples from the deployment set,
to a splittable representation;
the adversary that tries to identify, from one part of the split representation,
which dataset the given samples originated from;
and some kind of reconstruction loss
to ensure the splittable representation represents the input data well.
The following elements are in theory optional,
but are needed for the method to work with real-world data:
clustering and sampling to ensure the training batches are balanced in specific ways;
and subdivision of the batches into bags and the aggregation of the adversarial loss over the bags.
Roughly speaking, these elements strengthen the supervision signal for the invariance learning.

@@

% \section{Claims and contributions}%
\section{List of publications and author contributions}%
\label{sec:claims-contributions}
This thesis is based on 3 publications, corresponding to \rangechapref{ch:paper1}{ch:paper3}.
% The first one is concerned with \emph{label bias} and the other two with \emph{sampling bias}.
The following is a detailed listing of all the individual author contributions.

>>>>

The described method works particularly well with \acp{INN},
as they ensure that no information is lost that is unrelated to \(s\).
However, the price to pay for using \acp{INN} is higher memory requirement and slower training.
Thus, we also present a variant of the method using \iac{VAE},
which does not have the guarantee about preserving information,
but also does not suffer from the increased training cost as much.
\acp{VAE} are similar to \acp{INN}
in that their encoding conforms to a specific probability distribution,
from which we can sample our null-samples.
The choice between the two presented variants is determined
by whether the user is willing to accept higher training costs
for a lower probability of losing information needed for any prediction tasks.
However, the key element is simply any kind of encoder -- producing a split-encoding --
whose output can be subjected to adversarial training,
so encoders other than \acp{VAE} or \acp{INN} will potentially work as well.

@@

The absolutely essential elements for this method are the encoder (also referred to as `de-biaser')
that encodes both, samples from the training set and samples from the deployment set,
to a splittable representation;
the adversary that tries to identify, from one part of the split representation,
which dataset the given samples originated from;
and some kind of reconstruction loss
to ensure the splittable representation represents the input data well.
The following elements are in theory optional,
but are needed for the method to work with real-world data:
clustering and sampling to ensure that the training batches are balanced in specific ways;
subdivision of the batches into bags; and the aggregation of the adversarial loss over the bags.
Roughly speaking, these elements strengthen the supervision signal for the invariance learning.

@@

% \section{Claims and contributions}%
\section{List of publications and author contributions}%
\label{sec:claims-contributions}
This thesis is based on 3 publications (one of which is a work in progress),
corresponding to \rangechapref{ch:paper1}{ch:paper3}.
% The first one is concerned with \emph{label bias} and the other two with \emph{sampling bias}.
The following is a detailed listing of all the individual author contributions.

### 2021-09-06 ch/7-discussion.tex
Fix small errors found during corrections

% ========= unsupervised learning =======
One area of machine learning that has recently seen increased interest is unsupervised learning,
and the latter two chapters make use of it to some degree.
The exciting promise of unsupervised learning in general is
that labour-intensive labelling is not needed and so vast amounts of existing, unlabelled data can be put to good use.
One could ask the question whether bias-correcting methods are still needed, with access to so much data.
It could be that, while the data is certainly not perfect,
there is so much of it that the biased parts ``cancel each other out''.
However, recent investigations into the GPT models \citep{radford2018improving,radford2019language,brown2020language} do not seem to support this \citep{khalifa2021distributional}.
One reason for this might be the way these models are trained at the moment:
they maximise the probability assigned to the next token.
Thus, such a model has to account for the wide array of human opinions and assign a non-zero probability to all of them.
So, when asked to summarise a text \citep{stiennon2020learning}, GPT does not give the \emph{best} summary;
rather, it gives a summary that an average person might have written.
However, with the help of a very high-quality labelled dataset (that was expensive to create),
it was possible to fintetune GPT to actually produce very good summaries.
I suspect this pattern of learning the basics in an unsupervised fashion,
and then finetuning with high-quality labels, will continue in the future.
With these massive models,
it is, more than ever, crucial to build tools to make the biases within the models transparent.
Given the black-box nature of neural networks, this represents a major challenge.
% When examining these models trained on massive datasets through the lens of the dataset biases discussed in \chapref{ch:related-work},
% we encounter a novel problem:
% somewhere in the model,
% (an approximation of) the fundamental structures of reality are likely represented,
% but we cannot easily access this.
% ==============

>>>>

% ========= unsupervised learning =======
One area of machine learning that has recently seen increased interest is unsupervised learning,
and the latter two chapters make use of it to some degree.
The exciting promise of unsupervised learning in general is
that labour-intensive labelling is not needed and so vast amounts of existing, unlabelled data can be put to good use.
One could ask the question whether bias-correcting methods are still needed, with access to so much data.
It could be that, while the data is certainly not perfect,
there is so much of it that the biased parts ``cancel each other out''.
However, recent investigations into the GPT models \citep{radford2018improving,radford2019language,brown2020language} do not seem to support this \citep{khalifa2021distributional}.
One reason for this might be the way these models are trained at the moment:
they maximise the probability assigned to the next token.
Thus, such a model has to account for the wide array of human opinions and assign a non-zero probability to all of them.
So, when asked to summarise a text \citep{stiennon2020learning}, GPT does not give the \emph{best} summary;
rather, it gives a summary that an average person might have written.
However, with the help of a very high-quality labelled dataset (that was expensive to create),
it was possible to finetune GPT to actually produce very good summaries.
I suspect this pattern of learning the basics in an unsupervised fashion,
and then finetuning with high-quality labels, will continue in the future.
With these massive models,
it is, more than ever, crucial to build tools to make the biases within the models transparent.
Given the black-box nature of neural networks, this represents a major challenge.
% When examining these models trained on massive datasets through the lens of the dataset biases discussed in \chapref{ch:related-work},
% we encounter a novel problem:
% somewhere in the model,
% (an approximation of) the fundamental structures of reality are likely represented,
% but we cannot easily access this.
% ==============

### 2021-09-06 paper1/paper1.tex
Finish
Fix small errors found during corrections

% ============= content ====================
\textsc{Authors}:\;\;Thomas Kehrenberg$^1$, Zexun Chen$^{1,2}$, and Novi Quadrianto$^1$ \\
\textsc{Affiliations}:\\
$^1$ Predictive Analytics Lab (PAL), University of Sussex, Brighton, UK\\
$^2$ BioComplex Laboratory, University of Exeter, Exeter, UK \\
\textsc{Journal}:\;\;\textit{Frontiers in Artificial Intelligence}, Volume 3 \\
\textsc{DOI}:\;\;10.3389/frai.2020.00033 \\
\textsc{Note}:\;\;The appendix has been included as section~\ref{sec:tuning-appendix}.

@@

Furthermore, as currently formulated, the approach is only compatible with likelihood-based models
(which covers neural networks),
but the only real requirement is reasonably-well calibrated probabilities from the used model,
which are needed to compute the expected target labels.
Thus, future work could extend it to other kinds of models, like \acp{SVM},
for which solutions exists to transform the output to a probability;
the most common being Platt scaling \citep{platt1999probabilistic},
which applies a logistic transformation with two free parameters to the output of the \ac{SVM}.
The free parameters have to be estimated by iterating over the whole training set.
The remaining (substantial) challenge is then
to integrate the loss computed from these probabilities into the training of the \ac{SVM}.

>>>>

% ============= content ====================
\textsc{Authors}:\;\;Thomas Kehrenberg$^1$, Zexun Chen$^{1,2}$, and Novi Quadrianto$^1$ \\
\textsc{Affiliations}:\\
$^1$ Predictive Analytics Lab (PAL), University of Sussex, Brighton, UK\\
$^2$ BioComplex Laboratory, University of Exeter, Exeter, UK \\
\textsc{Journal}:\;\;\textit{Frontiers in Artificial Intelligence}, Volume 3 \\
\textsc{DOI}:\;\;\texttt{10.3389/frai.2020.00033} \\
\textsc{Note}:\;\;The appendix has been included as section~\ref{sec:tuning-appendix}.

@@

Furthermore, as currently formulated, the approach is only compatible with likelihood-based models
(which covers neural networks).
However, the only real requirement is reasonably-well calibrated probabilities from the used model,
which are needed to compute the expected target labels.
Thus, future work could extend it to other kinds of models, like \acp{SVM},
for which solutions exists to transform the output to a probability;
the most common being Platt scaling \citep{platt1999probabilistic},
which applies a logistic transformation with two free parameters to the output of the \ac{SVM}.
The free parameters have to be estimated by iterating over the whole training set.
The remaining (substantial) challenge is then
to integrate the loss computed from these probabilities into the training of the \ac{SVM}.

### 2021-09-06 paper3/paper3.tex
Fix small errors found during corrections

Our goal is that $z_y$ is invariant to the subgroup $s$.
However, what the adversarial loss actually enforces is the $z_y$ has the same distribution for the bags from the deployment set and the bags from the training set.
To ensure that the network learns the correct task,
it is crucial that subgroup membership is the only differing factor between the two types of bags.
The first step towards this goal is that bags from the deployment set are balanced (or \emph{perfect}):
all combinations of $s$ and $y$ appear at the same rate.
The second step is that, in the bags from the training set, the possible values of $y$ have to appear at the same rate,
because $y$ is meant to be preserved;
thus, $P(y_\mathit{tr}=0)=P(y_\mathit{tr}=1)=\mathellipsis$.
Finally, within the classes, subgroups should appear at equal rate:
$P(s_\mathit{tr}=0|y_\mathit{tr}=y')=P(s_\mathit{tr}=1|y_\mathit{tr}=y')=\mathellipsis$.
For example, if the bag size is 4, $Y$ and $S$ are binary, and the combination $(y=1,s=0)$ is missing,
then each bag should contain 2 samples of $(y=1,s=1)$ and 1 sample each of $(y=0,s=0)$ and $(y=0,s=1)$
(see also fig.~\ref{fig:matching-diagram}).
This ensures that the bags from the training set only differ in those samples from the deployment bags,
where a subgroup is missing.
%
% We know that the $y\perp s$ condition holds in the perfect bag, but not in the training set due to sampling bias.
% To do well, the discriminator should rely on this knowledge.
% More concretely, since the deployment and training set have differing support over $\Scal \times \Ycal$, namely $(\Scal_{tr} \times Y_{tr}) \subsetneq (\Scal_\mathit{perf} \times \Ycal_\mathit{perf})$, that support serves as an indicator of the distribution from which the data has been drawn.
% The scenarios we consider dictate $\Ycal_{tr} = \Ycal_\mathit{perf}$, making the disentangling well-posed.
% However, since we wish to use $\Scal_{dep} \times \Ycal_{dep} \setminus \Scal_{tr} \times \Ycal_{tr}$ as the training signal for the encoder, and not the relative frequency of the target classes,
% it is important that, like the deployment set, we weight the samples of the training set such that both $P(y_{tr})$ and $P(s_{tr}|y_{tr})$ are equal for all $s_{tr}, y_{tr} \in \Scal_{tr} \times \Ycal_{tr}$.
To guide the network towards the desired solution, we supplement this implicit constraint with the explicit constraint that $z_y$ be predictive of $y$, which we achieve using a linear predictor $l$. Whenever we have $\textrm{dim}(\mathcal{S}) > 1$ %(in our experiments this corresponds to the \emph{subgroup bias} setting)
we can also impose the same constraint on $z_s$, but with respect to $s$.
With these conditions met, to fool the discriminator,
the encoder must separate out information pertaining to $S$ into the space $z_s$ not part of the discriminator's input,
leaving only subgroup-unrelated information in $z_y$.

@@

Note that as long as the model sees the different \(s\)-\(y\)-combinations at the right proportions,
disentangling can (to an extent) also be achieved without the bag-wise loss
(\ie, when the bag size is set to 1).
The model then has to learn an implicit prior for which \(s\)-\(y\)-combinations to expect
in the two different sets -- training set and deployment set.
However, in our preliminary experiments we found this sample-wise approach to work much poorer
than using a bag-wise loss.
For those experiments, the batch was balanced in the same way as described above for the bags,
but the batch was not subdivided into bags
and there was no attention or aggregation method applied to the batch:
the loss was just computed per sample.
From the poor performance,
we inferred that the sample-wise loss lacks the necessary context
to easily differentiate between the two datasets.

@@

% \textbf{Disentangled representations via adversarial autoencoders}
%We want to learn a disentanglement using a perfect dataset as the supervision signal.
%
%To achieve this, our semi-supervised disentanglement learning framework (fig. \ref{fig:architecture} ) maps data points into independent factors of variations such that our training dataset $D_{\text{tr}}$ and our deployment set $D_c$ are indistinguishable; i.e., we would like transform $D_{\text{tr}}$ such that the independence condition $y\perp s$ holds.
%
%However, the deployment set itself $D_c$ might not amount to a suitable target distribution in which the independence condition holds.
%
%The idea is to make the deployment set $D_c$ approximately perfect by balancing it according to the cluster IDs, forming the set $D_p$.
%
% We create disentangled representations by training four separate neural networks, which we denote $f$, $g$, $h$ and $l$.
% %
% We use mean squared error for the reconstruction loss, and the binary cross entropy error for the discriminator loss.
% %
% For a dataset with mixed-typed attributes, we use a combination of mean squared error and (binary) cross entropy error for the reconstruction loss.
% %
% We optimise the proposed model using a scheduled update scheme where we freeze the weights of the autoencoder and predictor modules when we update the weights of the discriminator, and vice versa.
%
\section{Conclusion}
We have highlighted the problem that systematic bias can result in one or more subgroups having zero labelled data, and by doing so hope to have stimulated serious consideration for it (even if to be dismissed) when planning, building, evaluating and regulating machine learning systems.
We propose a two-step approach for addressing the resulting zero-label stratification problem.
First, we construct perfect bags from an unlabelled deployment set via  clustering. 
Second, we learn a disentangled representations using the perfect bags for adversarial distribution-matching. 
We empirically validate our framework on the Coloured MNIST, CelebA and Adult Income datasets, and find evidence that it is possible to maintain high performance on the subgroups with zero training data. 
We analyse our approach using the disentanglement calculus of \citet{ShuCheKumErmetal20} that relies on the notions of restrictiveness and consistency, and show that we can derive some guarantees.
We presented our approach in the context of biased data,
but it could be used as a general-purpose method for learning disentanglements
as long as the factors can be expressed by contrasting two data subsets;
that is, the data will be disentangled into two factors,
the first of which corresponding to the aspect that is common in the two sets,
and the second corresponding to the one that differs.
The method thus does not compete directly with \emph{unsupervised} disentanglement,
but is potentially more widely applicable than to only fairness problems.

>>>>

Our goal is that $z_y$ is invariant to the subgroup $s$.
However, what the adversarial loss actually enforces is that $z_y$ has the same distribution for the bags from the deployment set and the bags from the training set.
To ensure that the network learns the correct task,
it is crucial that subgroup membership is the only differing factor between the two types of bags.
The first step towards this goal is that bags from the deployment set are balanced (or \emph{perfect}):
all combinations of $s$ and $y$ appear at the same rate.
The second step is that, in the bags from the training set, the possible values of $y$ have to appear at the same rate,
because $y$ is meant to be preserved;
thus, $P(y_\mathit{tr}=0)=P(y_\mathit{tr}=1)=\mathellipsis$.
Finally, within the classes, subgroups should appear at equal rate:
$P(s_\mathit{tr}=0|y_\mathit{tr}=y')=P(s_\mathit{tr}=1|y_\mathit{tr}=y')=\mathellipsis$.
For example, if the bag size is 4, $Y$ and $S$ are binary, and the combination $(y=1,s=0)$ is missing,
then each bag should contain 2 samples of $(y=1,s=1)$ and 1 sample each of $(y=0,s=0)$ and $(y=0,s=1)$
(see also fig.~\ref{fig:matching-diagram}).
This ensures that the bags from the training set only differ in those samples from the deployment bags,
where a subgroup is missing.
%
% We know that the $y\perp s$ condition holds in the perfect bag, but not in the training set due to sampling bias.
% To do well, the discriminator should rely on this knowledge.
% More concretely, since the deployment and training set have differing support over $\Scal \times \Ycal$, namely $(\Scal_{tr} \times Y_{tr}) \subsetneq (\Scal_\mathit{perf} \times \Ycal_\mathit{perf})$, that support serves as an indicator of the distribution from which the data has been drawn.
% The scenarios we consider dictate $\Ycal_{tr} = \Ycal_\mathit{perf}$, making the disentangling well-posed.
% However, since we wish to use $\Scal_{dep} \times \Ycal_{dep} \setminus \Scal_{tr} \times \Ycal_{tr}$ as the training signal for the encoder, and not the relative frequency of the target classes,
% it is important that, like the deployment set, we weight the samples of the training set such that both $P(y_{tr})$ and $P(s_{tr}|y_{tr})$ are equal for all $s_{tr}, y_{tr} \in \Scal_{tr} \times \Ycal_{tr}$.
To guide the network towards the desired solution, we supplement this implicit constraint with the explicit constraint that $z_y$ be predictive of $y$, which we achieve using a linear predictor $l$. Whenever we have $\textrm{dim}(\mathcal{S}) > 1$ %(in our experiments this corresponds to the \emph{subgroup bias} setting)
we can also impose the same constraint on $z_s$, but with respect to $s$.
With these conditions met, to fool the discriminator,
the encoder must separate out information pertaining to $S$ into the space $z_s$ not part of the discriminator's input,
leaving only subgroup-unrelated information in $z_y$.

@@

Note that as long as the model sees the different \(s\)-\(y\)-combinations at the right proportions,
disentangling can (to an extent) also be achieved without the bag-wise loss
(\ie, when the bag size is set to 1).
The model then has to learn an implicit prior for which \(s\)-\(y\)-combinations to expect
in the two different sets -- training set and deployment set.
However, in our preliminary experiments we found this sample-wise approach to work much more poorly
than using a bag-wise loss.
For those experiments, the batch was balanced in the same way as described above for the bags,
but the batch was not subdivided into bags
and there was no attention or aggregation method applied to the batch:
the loss was just computed per sample.
From the poor performance,
we inferred that the sample-wise loss lacks the necessary context
to easily differentiate between the two datasets.

@@

% \textbf{Disentangled representations via adversarial autoencoders}
%We want to learn a disentanglement using a perfect dataset as the supervision signal.
%
%To achieve this, our semi-supervised disentanglement learning framework (fig. \ref{fig:architecture} ) maps data points into independent factors of variations such that our training dataset $D_{\text{tr}}$ and our deployment set $D_c$ are indistinguishable; i.e., we would like transform $D_{\text{tr}}$ such that the independence condition $y\perp s$ holds.
%
%However, the deployment set itself $D_c$ might not amount to a suitable target distribution in which the independence condition holds.
%
%The idea is to make the deployment set $D_c$ approximately perfect by balancing it according to the cluster IDs, forming the set $D_p$.
%
% We create disentangled representations by training four separate neural networks, which we denote $f$, $g$, $h$ and $l$.
% %
% We use mean squared error for the reconstruction loss, and the binary cross entropy error for the discriminator loss.
% %
% For a dataset with mixed-typed attributes, we use a combination of mean squared error and (binary) cross entropy error for the reconstruction loss.
% %
% We optimise the proposed model using a scheduled update scheme where we freeze the weights of the autoencoder and predictor modules when we update the weights of the discriminator, and vice versa.
%
\section{Conclusion}
We have highlighted the problem that systematic bias can result in one or more subgroups having zero labelled data, and by doing so hope to have stimulated serious consideration for it (even if to be dismissed) when planning, building, evaluating and regulating machine learning systems.
We propose a two-step approach for addressing the resulting zero-label stratification problem.
First, we construct perfect bags from an unlabelled deployment set via  clustering. 
Second, we learn a disentangled representations using the perfect bags for adversarial distribution-matching. 
We empirically validate our framework on the Coloured MNIST, CelebA and Adult Income datasets, and find evidence that it is possible to maintain high performance on the subgroups with zero training data. 
We analyse our approach using the disentanglement calculus of \citet{ShuCheKumErmetal20} that relies on the notions of restrictiveness and consistency, and show that we can derive some guarantees.
We presented our approach in the context of biased data,
but it could be used as a general-purpose method for learning disentanglements
as long as the factors can be expressed by contrasting two data subsets;
that is, the data will be disentangled into two factors,
the first of which corresponds to the aspect that is common in the two sets,
and the second corresponding to the one that differs.
The method thus does not compete directly with \emph{unsupervised} disentanglement,
but is potentially more widely applicable than to only fairness problems.
